Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2016-09-26 Paper/paper.tex
forgot to comment out the gray text

\begin{document}
\linenumbers

@@

\textcolor{gray}{%
Let us first delve into statistical errors. In
Fig.~\ref{fig:CIWidths90_Lambda_SNR}, we show how {\it precisely} can we
measure NSBH parameters $X=\{\mchirp,\eta,\chibh\}$ using BBH templates.
The three panels correspond to $\mchirp$ (top), $\eta$ (middle), and $\chibh$
(bottom), and show the width of these credible intervals $(\Delta X)^{90\%}$
as a function of BH mass/spin (within each sub-panel), and NS properties, i.e.
$\lambdans$ (downwards in each column)~\footnote{We restrict NS mass to
$1.35M_\odot$ and its spin to zero. Varying its tidal deformability $\lambdans$
does not significantly change the measurement uncertainties for non-tidal
binary parameters, as is evident from comparing the two rows in each panel of
Fig.~\ref{fig:CIWidths90_Lambda_SNR}.}.
% 
From the left-most column, we find that: (i) at $\rho=20, \mchirp$ is already
measured remarkably well - to a precision of $0.16\%$ of its true value, and
(ii) so is $\chibh$. (iii) Mass ratio $\eta$ is determined more loosely,
with $\gtrsim 25\%$ uncertainty. If the signal is even louder ($\rho\geq 30$),
all three measurements gain further precision, with $\eta$ errors shrinking down
to single-digit percents of its true value.
% % 
% We remind ourselves that this {\it precision}, however, is only
% meaningful so long as the measurement is {\it accurate} to begin with.
% We define $\arr_X$ as the ratio between systematic and statistical
% errors associated with the measurement of parameter $X$,
% \begin{equation}\label{eq:arr}
% \arr_X = \dfrac{(X^\mathrm{Median} - X^\mathrm{Injected})}{(\Delta X)^{90\%}},
% \end{equation}
% in order to compare the relative importance of both. Only when
% $|\arr_X| < 1$ can we hope that ignoring tidal effects in our templates
% will not hamper the measurement of non-tidal parameters from NSBH signals.
% Due to
% unmodeled tidal effects, parameter measurements performed on NSBH signals with
% BBH templates will have systematic biases. If these biases are too large,
% using non-tidal BBH templates should severely impair our inferencing ability.
% 
% there are physical effects that have not been
% incorporated in our (BBH) search templates which we use here to decipher NSBH
% signals. Therefore, we cannot know the {\it accuracy} of our parameter
% measurements unless we know how large the systematic biases due to an
% incomplete template model are. To quantify this, we find it useful to define
% $\arr_X$, the ratio of the systematic biases induced by unmodeled tidal effects in
% templates, to the statistical uncertainties associated with the measurement
% itself. If $\arr_X\gtrsim 1$, including the effects of NS tidal distortion
% (during inspiral) and possible disruption (close to merger) would be imperative
% to obtain {\it accurate} estimates of source masses and spins from observed GW
% signals.}
}

\textcolor{blue}{%
Credible intervals $(\Delta X)^{90\%}$ showing the precision with which
$X=\{\mchirp,\eta,\chibh\}$ can be measured, are presented in
Appendix~\ref{as1:nontidalerrors}.
}
We remind ourselves that this {\it precision} is only meaningful so long as the
measurement is {\it accurate} to begin with. Therefore, we define $\arr_X$ as
the ratio between systematic and statistical errors associated with the
measurement of parameter $X$,
\begin{equation}\label{eq:arr}
\arr_X = \dfrac{(X^\mathrm{Median} - X^\mathrm{Injected})}{(\Delta X)^{90\%}},
\end{equation}
in order to compare the relative magnitude of both. Only when
$|\arr_X| < 1$ can we ignore tidal effects in our templates
without hampering the measurement of non-tidal parameters from NSBH signals.
% 
We start with calculating $\arr_{\mchirp}$ as a function of various source
parameters and show it in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}.
$\mchirp$ is the leading order
mass combination that affects the GW strain emitted by compact binaries as they
spiral in, and is therefore determined the most precisely. We notice
immediately that for $\rho\leq 30$ the systematics are well under control
and so we can obtain reliable chirp mass estimates for NSBH signals using BBH
templates.
% 
For louder and less likely SNRs ($\rho\simeq 50$), we find that
$\arr_{\mchirp}$ can become comparable to unity, but only if the BH has
prograde spin $\chibh\gtrsim 0.4$, {\it and} the true NS tidal deformability
is large enough, s.t. $\lambdans \gtrsim 1000$.
% 
We can therefore conclude that only for very loud signals, with
$\rho\gtrsim 50-70$, will the inclusion of tidal terms in our template
models improve our $\mchirp$ estimation. For lower SNRs, inclusion of new
physical content in templates will instead get washed out by detector noise.
% 
In addition, we also note that $\arr_{\mchirp}\geq 0$ always,
i.e. $\mchirp$ is always being over-estimated. This is to be expected since
the tidal deformation and disruption of NSs close to merger reduces GW signal
power at high frequencies, making the resulting signal resemble a BBH signal of
lower merger frequency - and therefore of higher mass.

@@

In this section, we calculate the accuracy with which we measure $\lambdans$ from
{\it single} GW observations. We sample the same set of disruptive NSBH mergers
as in the previous section, i.e. those with $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000, 1500, 2000\}$;
fixing the NS mass $\mns=1.35 M_\odot$ and $\chins=0$. For each unique
combination of these parameters, we inject LEA+ signals into zero noise and
perform a fully Bayesian parameter estimation analysis of each with LEA+
templates. Our  priors on component masses and spins remain as in the
previous section, with mass-ratio additionally restricted to $2\leq q\leq 6$,
and $\lambdans$ sampled uniformly from $[0,4000]$.
% 
%
As an illustration of individual injections, we show the recovered probability
distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix
$q = \mbh /\mns = 5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, and
vary $\lambdans$ over $\{1000, 1500, 2000\}$ between the three panels.
The SNR is fixed at $\rho=50$. The darker shaded regions mark the $90\%$ credible
interval on $\lambdans$. We note that $\lambdans$ is estimated to within
$\pm 2000$ of its true value at this SNR. Another interesting thing to note 
is that while $(\Delta\lambdans)^{90\%}$ slowly grows with $\lambdans$, the
fractional uncertainty
\begin{equation}
\delta\lambdans^{90\%}:= (\Delta\lambdans)^{90\%}/\lambdans
\end{equation}
decreases instead.
%
\textcolor{blue}{%
Further illustrations, showing the correlation between tidal and non-tidal
parameters, are presented in Appendix~\ref{as1:illustrations}.
We will continue here to focus on the measurement of $\lambdans$ itself.
}
% 
\textcolor{gray}{%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
set $\rho=30$ (left panel) or $\rho=50$ (right panel).
%
We find that the measurement of $\lambdans$ is weakly degenerate with
other parameters, and at realistic SNRs it would improve by a few tens of 
percent if we knew non-tidal parameters to better accuracy. The predominant
factor that would enhance the measurement accuracy for $\lambdans$ is the
signal strength, and only when $\rho\gtrsim 50$ do we expect $\lambdans$ 
measurement to be limited by its degeneracy with non-tidal parameters at 
a factor of few level, as was reported by Ref.~\cite{Lackey:2013axa}.
}

>>>>

\begin{document}
%\linenumbers

@@

% \textcolor{gray}{%
% Let us first delve into statistical errors. In
% Fig.~\ref{fig:CIWidths90_Lambda_SNR}, we show how {\it precisely} can we
% measure NSBH parameters $X=\{\mchirp,\eta,\chibh\}$ using BBH templates.
% The three panels correspond to $\mchirp$ (top), $\eta$ (middle), and $\chibh$
% (bottom), and show the width of these credible intervals $(\Delta X)^{90\%}$
% as a function of BH mass/spin (within each sub-panel), and NS properties, i.e.
% $\lambdans$ (downwards in each column)~\footnote{We restrict NS mass to
% $1.35M_\odot$ and its spin to zero. Varying its tidal deformability $\lambdans$
% does not significantly change the measurement uncertainties for non-tidal
% binary parameters, as is evident from comparing the two rows in each panel of
% Fig.~\ref{fig:CIWidths90_Lambda_SNR}.}.
% % 
% From the left-most column, we find that: (i) at $\rho=20, \mchirp$ is already
% measured remarkably well - to a precision of $0.16\%$ of its true value, and
% (ii) so is $\chibh$. (iii) Mass ratio $\eta$ is determined more loosely,
% with $\gtrsim 25\%$ uncertainty. If the signal is even louder ($\rho\geq 30$),
% all three measurements gain further precision, with $\eta$ errors shrinking down
% to single-digit percents of its true value.
% % % 
% % We remind ourselves that this {\it precision}, however, is only
% % meaningful so long as the measurement is {\it accurate} to begin with.
% % We define $\arr_X$ as the ratio between systematic and statistical
% % errors associated with the measurement of parameter $X$,
% % \begin{equation}\label{eq:arr}
% % \arr_X = \dfrac{(X^\mathrm{Median} - X^\mathrm{Injected})}{(\Delta X)^{90\%}},
% % \end{equation}
% % in order to compare the relative importance of both. Only when
% % $|\arr_X| < 1$ can we hope that ignoring tidal effects in our templates
% % will not hamper the measurement of non-tidal parameters from NSBH signals.
% % Due to
% % unmodeled tidal effects, parameter measurements performed on NSBH signals with
% % BBH templates will have systematic biases. If these biases are too large,
% % using non-tidal BBH templates should severely impair our inferencing ability.
% % 
% % there are physical effects that have not been
% % incorporated in our (BBH) search templates which we use here to decipher NSBH
% % signals. Therefore, we cannot know the {\it accuracy} of our parameter
% % measurements unless we know how large the systematic biases due to an
% % incomplete template model are. To quantify this, we find it useful to define
% % $\arr_X$, the ratio of the systematic biases induced by unmodeled tidal effects in
% % templates, to the statistical uncertainties associated with the measurement
% % itself. If $\arr_X\gtrsim 1$, including the effects of NS tidal distortion
% % (during inspiral) and possible disruption (close to merger) would be imperative
% % to obtain {\it accurate} estimates of source masses and spins from observed GW
% % signals.}
% }

%\textcolor{blue}{%
Credible intervals $(\Delta X)^{90\%}$ showing the precision with which
$X=\{\mchirp,\eta,\chibh\}$ can be measured, are presented in
Appendix~\ref{as1:nontidalerrors}.
%}
We remind ourselves that this {\it precision} is only meaningful so long as the
measurement is {\it accurate} to begin with. Therefore, we define $\arr_X$ as
the ratio between systematic and statistical errors associated with the
measurement of parameter $X$,
\begin{equation}\label{eq:arr}
\arr_X = \dfrac{(X^\mathrm{Median} - X^\mathrm{Injected})}{(\Delta X)^{90\%}},
\end{equation}
in order to compare the relative magnitude of both. Only when
$|\arr_X| < 1$ can we ignore tidal effects in our templates
without hampering the measurement of non-tidal parameters from NSBH signals.
% 
We start with calculating $\arr_{\mchirp}$ as a function of various source
parameters and show it in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}.
$\mchirp$ is the leading order
mass combination that affects the GW strain emitted by compact binaries as they
spiral in, and is therefore determined the most precisely. We notice
immediately that for $\rho\leq 30$ the systematics are well under control
and so we can obtain reliable chirp mass estimates for NSBH signals using BBH
templates.
% 
For louder and less likely SNRs ($\rho\simeq 50$), we find that
$\arr_{\mchirp}$ can become comparable to unity, but only if the BH has
prograde spin $\chibh\gtrsim 0.4$, {\it and} the true NS tidal deformability
is large enough, s.t. $\lambdans \gtrsim 1000$.
% 
We can therefore conclude that only for very loud signals, with
$\rho\gtrsim 50-70$, will the inclusion of tidal terms in our template
models improve our $\mchirp$ estimation. For lower SNRs, inclusion of new
physical content in templates will instead get washed out by detector noise.
% 
In addition, we also note that $\arr_{\mchirp}\geq 0$ always,
i.e. $\mchirp$ is always being over-estimated. This is to be expected since
the tidal deformation and disruption of NSs close to merger reduces GW signal
power at high frequencies, making the resulting signal resemble a BBH signal of
lower merger frequency - and therefore of higher mass.

@@

In this section, we calculate the accuracy with which we measure $\lambdans$ from
{\it single} GW observations. We sample the same set of disruptive NSBH mergers
as in the previous section, i.e. those with $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000, 1500, 2000\}$;
fixing the NS mass $\mns=1.35 M_\odot$ and $\chins=0$. For each unique
combination of these parameters, we inject LEA+ signals into zero noise and
perform a fully Bayesian parameter estimation analysis of each with LEA+
templates. Our  priors on component masses and spins remain as in the
previous section, with mass-ratio additionally restricted to $2\leq q\leq 6$,
and $\lambdans$ sampled uniformly from $[0,4000]$.
% 
%
As an illustration of individual injections, we show the recovered probability
distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix
$q = \mbh /\mns = 5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, and
vary $\lambdans$ over $\{1000, 1500, 2000\}$ between the three panels.
The SNR is fixed at $\rho=50$. The darker shaded regions mark the $90\%$ credible
interval on $\lambdans$. We note that $\lambdans$ is estimated to within
$\pm 2000$ of its true value at this SNR. Another interesting thing to note 
is that while $(\Delta\lambdans)^{90\%}$ slowly grows with $\lambdans$, the
fractional uncertainty
\begin{equation}
\delta\lambdans^{90\%}:= (\Delta\lambdans)^{90\%}/\lambdans
\end{equation}
decreases instead.
%
% \textcolor{blue}{%
Further illustrations, showing the correlation between tidal and non-tidal
parameters, are presented in Appendix~\ref{as1:illustrations}.
We will continue here to focus on the measurement of $\lambdans$ itself.
% }
% 
% \textcolor{gray}{%
% Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
% mass, spin, and tidal parameter measurements. We keep the binary parameters
% as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
% set $\rho=30$ (left panel) or $\rho=50$ (right panel).
% %
% We find that the measurement of $\lambdans$ is weakly degenerate with
% other parameters, and at realistic SNRs it would improve by a few tens of 
% percent if we knew non-tidal parameters to better accuracy. The predominant
% factor that would enhance the measurement accuracy for $\lambdans$ is the
% signal strength, and only when $\rho\gtrsim 50$ do we expect $\lambdans$ 
% measurement to be limited by its degeneracy with non-tidal parameters at 
% a factor of few level, as was reported by Ref.~\cite{Lackey:2013axa}.
% }
